Predicting Erectile Function in Patients Undergoing Stereotactic Body Radiation Therapy

### 2017-01-26 paper_draft.Rmd
cv
spell check

## Abstract
Stereotactic Body Radiation Therapy (SBRT) is a new method for treating prostate cancer. This study looks at the potential for SBRT patients to experience erectile dysfunction, an important side-effect that is associated with the treatment. Using a logistic regression model, this study finds that baseline variables including age, score on a quality of life survey, erectile function at baseline, and prostate-specific antigen levels can be used to predict the probability of having erectile function two years after recieving SBRT. The model performs well, with a value of 0.73 area under curve (AUC). Following the results from the model, we conclude that around 38 percent of patients who undergo SBRT are likely to have erectile function at two years, compared to 49 percent at baseline.

@@

To check for outliers or distribution oddities, we plotted a histogram for each of the continuous variables. If we found outliers in a histogram, we would first look at those outliers to determine if their values are theoretically possible. If the outliers were clearly typos, we would contact the data provider and then remove them. Any values that were strange but theoretically possible were left in the model. In particuar, subject 67 has a BMI of 112, which is very different from the other BMI values but still within the realm of possibility.

@@

To analyze the possible multicolinear effects of including both HRQOL and Erection function at baseline variables in the model, we fit the model with each variable individually, and then with both. We decided to keep both variables in the model because both variables were shown to be highly significant when fit toggether, and because models without both variables performed worse.

@@

The ROC curve produced from our model can be found in Figure 1. The area under the ROC curve is 0.842, indicating reasonable model performance. Using the ROC curve, we used the minimum Eucliden distance between the ROC curve and the upper left corner of the graph window to calculate an optimal classification threshold of 0.426. Using this threshold, 38 percent of patients in the study were predicted to have erectile function after two years down from 49 percent at baseline, which represents a 0.225 error rate from the actual results in the data. For comparison, using 0.5 as the cutoff, 30 percent of patients were expected to retain erectile function, representing a 0.206 error rate. In actuality, 31 percent of patients had function after two years. The benefit of using the calculated threshold of 0.426 over 0.5 is it is more optimistic--by classifying more people as retaining erectile function, we increase the true negative rate, and are less likely to scare off patients from opting for the treatment.

>>>>

## Abstract
Stereotactic Body Radiation Therapy (SBRT) is a new method for treating prostate cancer. This study looks at the potential for SBRT patients to experience erectile dysfunction, an important side-effect that is associated with the treatment. Using a logistic regression model, this study finds that baseline variables including age, score on a quality of life survey, erectile function at baseline, and prostate-specific antigen levels can be used to predict the probability of having erectile function two years after receiving SBRT. The model performs well, with a value of 0.73 area under curve (AUC). Following the results from the model, we conclude that around 38 percent of patients who undergo SBRT are likely to have erectile function at two years, compared to 49 percent at baseline.

@@

To check for outliers or distribution oddities, we plotted a histogram for each of the continuous variables. If we found outliers in a histogram, we would first look at those outliers to determine if their values are theoretically possible. If the outliers were clearly typos, we would contact the data provider and then remove them. Any values that were strange but theoretically possible were left in the model. In particular, subject 67 has a BMI of 112, which is very different from the other BMI values but still within the realm of possibility.

@@

To analyze the possible multicollinear effects of including both HRQOL and Erectile function at baseline variables in the model, we fit the model with each variable individually, and then with both. We decided to keep both variables in the model because both variables were shown to be highly significant when fit together, and because models without both variables performed worse.

@@

The ROC curve produced from our model can be found in Figure 1. The area under the ROC curve is 0.842, indicating reasonable model performance. Using the ROC curve, we used the minimum Euclidean distance between the ROC curve and the upper left corner of the graph window to calculate an optimal classification threshold of 0.426. Using this threshold, 38 percent of patients in the study were predicted to have erectile function after two years down from 49 percent at baseline, which represents a 0.225 error rate from the actual results in the data. For comparison, using 0.5 as the cutoff, 30 percent of patients were expected to retain erectile function, representing a 0.206 error rate. In actuality, 31 percent of patients had function after two years. The benefit of using the calculated threshold of 0.426 over 0.5 is it is more optimistic--by classifying more people as retaining erectile function, we increase the true negative rate, and are less likely to scare off patients from opting for the treatment. We investigated the calculated error rate by using 10-fold cross validation, and found the estimated error rate to be around 0.24 at both 0.426 and 0.5 thresholds.
